---
title: "R class - Part 2"
output: html_notebook
---

### Load the tidyverse, airbnb data, and climate data

```r
library(tidyverse)
airbnb <- read_csv('https://bit.ly/3oadz2L')
climate <- read_csv('https://bit.ly/3kKErEb')
```


### Save data locally

```{r}
write_csv(x = airbnb, file = 'airbnb.csv')
# save your data locally using %>% 
# airbnb %>% write_csv(file = 'airbnb.csv')
```

### Working Directory

Use`getwd()` to see where your working directory path and `setwd()` to change your working directory

```{r}
getwd()
setwd('/Users/grahambearden/projects/r_course_evans_school/')
```

### Solution to the troubleshooting exercise

```{r}
library(tidyverse)
airbnb <- read_csv('https://bit.ly/3oadz2L')
reviews <- airbnb$reviews
price <- airbnb$price
med_reviews <- median(reviews)
max_price <- max(price)
med_reviews
max_price
```

### Create a new data object called `airbnb_new`

```{r}
airbnb_new <- airbnb %>% 
  select(host_id, reviews, rating, price, accommodates) %>% 
  mutate(
    cost_per_person = price / accommodates,
    mean_price = mean(price)
    )

airbnb_new
```

### `%in%` is really useful! 

You can it to search for a single value in a longer vector.

```{r}
city_1 <- 'San Francisco'
city_2 <- 'Portland'
california <- c('San Francisco', 'Los Angeles', 'San Diego')
ifelse(! city_2 %in% california, 'california city', 'non-california')

airbnb %>% 
  select(host_id, address, reviews, rating, price, accommodates) %>% 
  mutate(
    cost_per_room = accommodates / price,
    mean_price = mean(price),
    seattle_address = ifelse(address == 'Seattle', 1, 0),
    large = ifelse(accommodates > 5, 1, 0)
    ) 

```

```{r}
high_uncertainty <- climate %>% 
  mutate(
    high_uncertainty = ifelse(uncertainty > 1.5, 1, 0),
    winter = ifelse(month %in% c(1, 2, 12), 1, 0)
    ) %>%
  select(country, city, uncertainty, high_uncertainty, month, winter)
```

### Calculate % of rentals in Seattle

```{r}
num_seattle_address <- airbnb %>% 
  mutate(seattle_address = ifelse(address == 'Seattle', 1, 0)) %>% 
  pull(seattle_address) %>% 
  sum()

num_obs <- nrow(airbnb)

num_seattle_address / num_obs # percent of rentals in Seattle
```

### `transmute()` combines `mutate()`  and `select()`

Almost always when I see `mutate()` and `select()` in my code, I replace it with `transmute()`.

```{r}
airbnb %>% 
  transmute(
    city = address,
    cost = price,
    description = name
    )
```

```{r}
climate %>% 
  transmute(
    year,
    location = city,
    temp = ifelse(is.na(temp), 0, temp)
    )
```

### You can use more than one condition when subsetting your data.

Use `&` and `|` for multiple conditions

```{r}
airbnb %>% 
  filter(price > 50 & price < 100 & room_type == 'Entire home/apt') %>% 
  select(room_id, host_id, address, room_type, rating, name, price)
```

The code above and below do the same thing.

```{r}
airbnb %>% 
  filter(price > 50) %>% 
  filter(price < 100) %>% 
  filter(room_type == 'Entire home/apt') %>% 
  select(room_id, host_id, address, room_type, rating, name, price)
```