---
pagetitle: "Analysis Memo"
output:
  html_document:
    theme: null
    highlight: null
    mathjax: null
    css: memo.css
params:
  to: "Director, Earth Institute"
  from: "Your Name, Climate Analyst"
  date: !r format(Sys.Date(), "%B %d, %Y")
  re: "Warming in Brazil's largest cities, 1951-2012"
---

```{r setup, include = FALSE}
library(tidyverse)
library(scales)

# Styles live in memo.css; the header layout lives in memo_header.html;
# figure options (wrap, wrap.width) are explained in memo_setup.R
source("memo_setup.R")
```

```{r data}
# Replace this chunk with your own data
climate <- read_csv("https://bit.ly/3kKErEb", show_col_types = FALSE)

brazil <- climate |>
  filter(country == "Brazil", year >= 1951, year <= 2012, !is.na(temp)) |>
  mutate(period = if_else(year <= 1980, "1951-1980", "1983-2012")) |>
  filter(year <= 1980 | year >= 1983)

change <- brazil |>
  group_by(city, period) |>
  summarise(mean_temp = mean(temp), .groups = "drop") |>
  pivot_wider(names_from = period, values_from = mean_temp) |>
  mutate(change = `1983-2012` - `1951-1980`) |>
  arrange(desc(change))

fastest <- slice_max(change, change, n = 1)
slowest <- slice_min(change, change, n = 1)
```

```{r header}
memo_header(params)
```

<!-- Opening summary: the bottom line in two or three sentences. -->

**Average temperatures rose in all `r nrow(change)` of Brazil's largest cities between 1951-1980 and 1983-2012. `r fastest$city` warmed the most (`r number(fastest$change, accuracy = 0.01)`°C) and `r slowest$city` warmed the least (`r number(slowest$change, accuracy = 0.01)`°C). [Replace with your recommendation in one sentence.]**

## Background

[Describe the question, who is asking, and why it matters now. Cite sources in footnotes.^[Author, *Title*, Publisher, Year, URL.]]

## Findings

<!-- Repeat this Finding / Evidence / Implication block for each finding. -->

### Finding 1: Every city is warmer than it was a generation ago

```{r trend-plot, wrap = "left", wrap.width = 0.5, fig.width = 5, fig.height = 3.6}
brazil |>
  group_by(city, year) |>
  summarise(temp = mean(temp), .groups = "drop") |>
  ggplot(aes(x = year, y = temp, color = city)) +
  geom_line(linewidth = 0.6) +
  labs(x = NULL, y = "Mean temperature (°C)", color = NULL,
       title = "Annual mean temperature, 1951-2012") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title.position = "plot") +
  guides(color = guide_legend(nrow = 2))
```

**Evidence:** Each line shows a city's average temperature by year. The cities differ by several degrees, but they move together, and each one ends the period warmer than it started. Year-to-year swings are large, so comparing 30-year averages gives a steadier picture than comparing single years. [Replace with the evidence for your finding.]

**Implication:** [Explain what this finding means for the reader's decision. Text in this section wraps around the figure. Change `wrap = "left"` to `wrap = "right"` to move the figure, or change `wrap.width` to resize it. Keep this text at least as tall as the figure, or the next heading will start below it.]

### Finding 2: Warming was uneven across cities

**Evidence:** The change in average temperature ranged from `r number(slowest$change, accuracy = 0.01)`°C in `r slowest$city` to `r number(fastest$change, accuracy = 0.01)`°C in `r fastest$city`. [Replace with the evidence for your finding.]

```{r change-plot, fig.align = "center", out.width = "75%", fig.width = 6, fig.height = 3}
change |>
  ggplot(aes(x = change, y = reorder(city, change))) +
  geom_segment(aes(x = 0, xend = change, yend = city), color = "grey70") +
  geom_point(size = 3, color = "#B5532A") +
  labs(x = "Change in mean temperature (°C), 1951-1980 to 1983-2012", y = NULL,
       title = "Change in average temperature by city") +
  theme_minimal(base_size = 12) +
  theme(plot.title.position = "plot")
```

**Implication:** [Explain what this finding means for the reader's decision. The figure above is centered on its own line. Change `fig.align` to `"left"` or `"right"` to align it, and `out.width` to resize it.]

## Summary Table

```{r change-table}
change |>
  transmute(
    City = city,
    `1951-1980 (°C)` = number(`1951-1980`, accuracy = 0.1),
    `1983-2012 (°C)` = number(`1983-2012`, accuracy = 0.1),
    `Change (°C)` = number(change, accuracy = 0.01)
  ) |>
  kable(caption = "Average monthly temperature by period")
```

## Recommendation

[State your recommendation, the main trade-off, and the next step.]

## Data and Methods

Temperatures come from Berkeley Earth's land temperature records for major cities.^[Berkeley Earth, *Climate Change: Earth Surface Temperature Data*, Kaggle, https://www.kaggle.com/datasets/berkeleyearth/climate-change-earth-surface-temperature-data.] Each period averages the monthly mean temperatures for each city across 30 years. [Note any limits of the data the reader should know about.]
