---
pagetitle: "Decision Memo"
output:
  html_document:
    theme: null
    highlight: null
    mathjax: null
    css: memo.css
params:
  to: "Councilmember, Seattle City Council"
  from: "Your Name, Office of Planning and Community Development"
  date: !r format(Sys.Date(), "%B %d, %Y")
  re: "Short-term rental listings in the Seattle area"
---

```{r setup, include = FALSE}
library(tidyverse)
library(scales)

# Styles live in memo.css; the header layout lives in memo_header.html;
# figure options (wrap, wrap.width) are explained in memo_setup.R
source("memo_setup.R")
```

```{r data}
# Replace this chunk with your own data
airbnb <- read_csv("https://bit.ly/3oadz2L", show_col_types = FALSE)

n_listings    <- nrow(airbnb)
share_seattle <- mean(airbnb$address == "Seattle")
share_entire  <- mean(airbnb$room_type == "Entire home/apt")

by_city <- airbnb |>
  group_by(address) |>
  summarise(
    listings     = n(),
    share_entire = mean(room_type == "Entire home/apt"),
    median_price = median(price, na.rm = TRUE)
  )
city <- function(x) filter(by_city, address == x)
```

```{r header}
memo_header(params)
```

<!-- Opening summary: state the bottom line in two or three sentences. A busy
reader should be able to stop here and still know what you found and what you
recommend. -->

**There are `r comma(n_listings, accuracy = 1)` Airbnb listings in Seattle, Bellevue, Kirkland, and Redmond, and `r percent(share_seattle, accuracy = 1)` of them are in Seattle. Most listings (`r percent(share_entire, accuracy = 1)`) rent out an entire home or apartment rather than a room. [Replace with your recommendation in one sentence.]**

## Background

<!-- Why is this memo landing on the reader's desk now? Keep it to the
context they need to understand the analysis. -->

[Describe the policy question and why it matters to the reader. Cite sources in footnotes.^[Author, *Title*, Publisher, Year, URL.]]

## Analysis

```{r listings-plot, wrap = "right", wrap.width = 0.45, fig.width = 4, fig.height = 2.8}
airbnb |>
  count(address, room_type) |>
  ggplot(aes(x = n, y = reorder(address, n, sum), fill = room_type)) +
  geom_col() +
  scale_x_continuous(labels = comma) +
  labs(x = "Listings", y = NULL, fill = NULL, title = "Listings by city and room type") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", plot.title.position = "plot")
```

<!-- Lead with the finding, then give the evidence. This paragraph wraps
around the figure: change wrap = "right" to wrap = "left" to move the figure,
or change wrap.width to resize it. Keep the paragraph at least as tall as the
figure, or the text after it will wrap too. -->

Short-term rentals are concentrated in Seattle. The city has `r comma(city("Seattle")$listings)` listings, more than `r floor(city("Seattle")$listings / sum(by_city$listings[by_city$address != "Seattle"]))` times as many as Bellevue, Kirkland, and Redmond combined. The type of rental also differs across the region. In Seattle, `r percent(city("Seattle")$share_entire, accuracy = 1)` of listings rent out an entire home or apartment, compared with `r percent(city("Bellevue")$share_entire, accuracy = 1)` in Bellevue and `r percent(city("Redmond")$share_entire, accuracy = 1)` in Redmond, where renting out a private room is more common. Whole-home rentals matter most for housing policy because they can take units out of the long-term rental market. [Replace with what the evidence means for the reader's decision, and note any limits of the data, such as when it was collected.]

```{r price-table}
airbnb |>
  group_by(City = address) |>
  summarise(
    Listings = n(),
    `Entire home (%)` = round(100 * mean(room_type == "Entire home/apt")),
    `Median nightly price ($)` = median(price, na.rm = TRUE)
  ) |>
  arrange(desc(Listings)) |>
  kable(format.args = list(big.mark = ","), caption = "Listings and prices by city")
```

## Recommendation

<!-- Tell the reader what to do, and what happens next. -->

[State your recommendation, the main trade-off, and the next step.]
