---
title: "Evans School R Class - Part 1"
output: html_notebook
---

# My notes
## Create a title with hashes!
#### The more hashes, the smaller the title

I'm going to write notes here.

Sometimes I like to *italicize*. Other times I like put something in **bold**.

I can also turn plain text into code with single tic marks (`foo <- 1:5`)

### Install packages

You only need to install a package once. So no need to run `install.packages('tidyverse')` again. But save or remember this code for install other packages.

```{r}
install.packages('tidyverse')
```

### Load libraries at the top of your script

`tidyverse` is an ensemble library, which means that it is a collection of libraries. When you load it, you actually load many. 

```{r}
library(tidyverse)
```

### Summary statistics example

```{r}
some_numbers <- c(1,2,NA, 4)
min(some_numbers, na.rm = TRUE)
mean(some_numbers)
max(some_numbers)
sd(some_numbers)
```


### Arguments in functions.

A `cat_function()` for cat lovers. In the case of `cat_function()`, `love` if the argument. Depending on what `love` equals determines the output.

#### Here you create the function

```{r}
cat_function <- function(love = TRUE){
  # SOURCE: https://hilaryparker.com/category/data-analysis/
  if(love == TRUE){
      print('I love cats!')
  }
  else {
      print('I am not a cool person.') 
  }
}
```

#### And here you run the function

```{r}
cat_function(love = FALSE)
```


### Use class to determine a variable type

```{r}
class(c('Washington', 'Oregon', 'Idaho'))
is.character(c('Washington', 'Oregon', 'Idaho'))
is.factor(c('Washington', 'Oregon', 'Idaho'))
```

### Use the assignment operator to create a data object (or variable)

```{r}
pnw <- c('Washington', 'Oregon', 'Idaho')
pnw_factor <- as.factor(pnw)
```

### Example of a tibble

```{r}
tibble(
  x = c(1:3)
  , y = c(4:6)
  , z = c('Washington', 'Oregon', 'Idaho')
  )
```

### "Grab" variables from a tibble (or data frame) with $

```{r}
cars$speed
```

### Import data with read_csv()

The bit.ly url routes to a csv file on Github.

```{r}
airbnb <- read_csv('https://bit.ly/3oadz2L')
```

...but maybe you want to save the data to a local drive and import it from there...

```{r}
# Save data to a local drive
# Yes, you can write comments within the code chunk too
write_csv(airbnb, 'airbnb.csv')

# Import data from the local drive
airbnb_local <- read_csv('airbnb.csv')

# Wondering where you saved the airbnb data?
getwd()

# Want to change the drive to where you save?
setwd()
```

#### Summary is a very useful way to look at your data a first time

```{r}
summary(airbnb)
```

#### Nesting functions is possible and can be useful

```{r}
sort(unique(airbnb$rating))
length(unique(airbnb$room_type))
```

### Use help to learn about a specific function

```{r}
help(lm)
help(is.character)
help(min)
```

